---
title: "The Most Harmful Weather Events Across the United States"
---

# The Most Harmful Weather Events Across the United States

## Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.
The basic goal of this investigation is to explore the NOAA Storm Database and answer some basic questions about severe weather events:
* Across the United States, which types of events are most harmful with respect to population health?
* Across the United States, which types of events have the greatest economic consequences?

## Data Processing

### Data set

The U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database is used for investigation. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. The data is downloaded in the form of a comma-separated-value file compressed via the bzip2 algorithm from [source](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).
There is also some documentation of the database available:
* [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)
The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

### Data download

As mentioned before, data is downloaded from [source](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) and placed in 'data/' directory. 'dplyr' package is used immediatelly for data processing.

```{r cache = TRUE}
if (!file.exists("data")) {
	dir.create("data")
}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", dest = "data/storm_data.csv.bz2")

library(dplyr)
data <- tbl_df(read.csv("data/storm_data.csv.bz2"))
```

Let's verify data that was downloaded:
```{r}
data
```
```{r}
summary(data)
```

### Data preprocessing

Mostly, we are interested in event types, population and economical damage that was caused by them. Following columns contain raw observations that we are interested in:
* EVTYPE - event type
* FATALITIES, INJURIES - population damage
* PROPDMG, PROPDMGEXP - property damage
* CROPDMG, CROPDMGEXP - crop damage

The data they contain is not ready for processing, so we need to prepare it, one by one.

#### EVTYPE

Let's exam EVTYPE column:

```{r}
unique(data$EVTYPE)
length(unique(data$EVTYPE))
```

As we can see, this column contains some irreleval values, mistakes, etc. Size of error them introduce is not huge, so we will perform only some basic cleaning:
```{r}
data <- mutate(data, EVTYPE = toupper(EVTYPE))
length(unique(data$EVTYPE))
```

#### FATALITIES, INJURIES

Let's exam FATALITIES and INJURIES columns:
```{r}
summary(data$FATALITIES)
summary(data$INJURIES)
```

As we can see, they are ready for processing.

#### PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP

PROPDMG/CROPDMG describes value of damages causes and PROPDMGEXP/CROPDMGEXP describes magnitude as following:
* K for $1000
* M for $1000000
* B for $1000000000

So we need to merge them into single variable to perform effective furter analysys.

```{r}
data <- mutate(data, 
	PROPERTY_DAMAGE = PROPDMG * ifelse(PROPDMGEXP == 'B', 1000000000, ifelse(PROPDMGEXP == 'M', 1000000, ifelse(PROPDMGEXP == 'K', 1000, 1))), 
	CROP_DAMAGE = CROPDMG * ifelse(CROPDMGEXP == 'B', 1000000000, ifelse(CROPDMGEXP == 'M', 1000000, ifelse(CROPDMGEXP == 'K', 1000, 1)))
)
```
```{r}
summary(data$PROPERTY_DAMAGE)
summary(data$CROP_DAMAGE)
```

#### Per event damages

Using dplyr grouping, let's group different measurements of damage by their event types:

```{r}
data_per_type <- summarise(group_by(data, EVTYPE), 
	TOTAL_FATALITIES = sum(FATALITIES), 
	TOTAL_INJURIES = sum(INJURIES),
	TOTAL_PROPERTY_DAMAGE = sum(PROPERTY_DAMAGE),
	TOTAL_CROP_DAMAGE = sum(CROP_DAMAGE)
)
data_per_type
```

#### Total value

In the end, to simplify further processing, let's calculate some summary damage measurements. Heuristically, we will think about fatality 5 times worse as injury and property damage equal to crop:

```{r}
data_per_type <- mutate(data_per_type,
	TOTAL_POPULATION_DAMAGE = 5 * TOTAL_FATALITIES + TOTAL_INJURIES,
	TOTAL_ECONOMICAL_DAMAGE = TOTAL_PROPERTY_DAMAGE + TOTAL_CROP_DAMAGE
)
data_per_type
```

## Results

### Population damage

Let's first look at top 10 event types that causes fatalities:

```{r}
ggplot(
	slice(arrange(data_per_type, desc(TOTAL_FATALITIES)), 1:10), 
	aes(x = reorder(EVTYPE, TOTAL_FATALITIES), y = TOTAL_FATALITIES)
) + 
geom_bar(stat="identity") + 
coord_flip() + 
labs(x = 'Event type') +
labs(y = 'Fatalities number') +
labs(title = 'Total fatalities number caused by top 10 event types')
```

Or, more detailed, table with top 20 event types:

```{r}
select(slice(arrange(data_per_type, desc(TOTAL_FATALITIES)), 1:20), EVTYPE, TOTAL_FATALITIES)
```

Now, let's look at top 10 event types that causes injuries:

```{r}
ggplot(
	slice(arrange(data_per_type, desc(TOTAL_INJURIES)), 1:10), 
	aes(x = reorder(EVTYPE, TOTAL_INJURIES), y = TOTAL_INJURIES)
) + 
geom_bar(stat="identity") + 
coord_flip() + 
labs(x = 'Event type') +
labs(y = 'Injuries number') +
labs(title = 'Total injuries number caused by top 10 event types')
```

Or, more detailed, table with top 20 event types:

```{r}
select(slice(arrange(data_per_type, desc(TOTAL_INJURIES)), 1:20), EVTYPE, TOTAL_INJURIES)
```

Finally, let's look at top 20 event types, calculated by our heuristic algo (fatality 5 times worse as injury):

```{r}
select(slice(arrange(data_per_type, desc(TOTAL_POPULATION_DAMAGE)), 1:20), EVTYPE, TOTAL_FATALITIES, TOTAL_INJURIES)
```

```{r}
ggplot(
	slice(arrange(data_per_type, desc(TOTAL_ECONOMICAL_DAMAGE)), 1:10), 
	aes(x = reorder(EVTYPE, TOTAL_ECONOMICAL_DAMAGE), y = TOTAL_ECONOMICAL_DAMAGE)
) + 
geom_bar(stat="identity") + 
coord_flip()
```